A Generic and Efficient E-field Parallel Imaging Correlator for Next-Generation Radio Telescopes

### 2015-10-23 imaging_demo/ms.tex
Add more text and figures in sections 3 and 4
From a3bda8c (master):
=====================

In epic.bib,

(*) Rearrange references alphabetically
(*) Add Beardsley et al. (2012) to references

In ms.tex,

(*) Add more text and figures in sections 3 and 4

The antenna array aperture illumination over the entire grid,
$\widetilde{W}(\mathbf{r})$, is obtained by a projection of the individual
antenna aperture illuminations:
\begin{align}
  \widetilde{W}(\mathbf{r}) &= \sum_a \widetilde{W}_a(\mathbf{r}-\mathbf{r}_a) \\
                            &= \mathbf{M}(\mathbf{r},\mathbf{e})\,\mathcal{I}(\mathbf{e}),
\end{align}
where, $\mathcal{I}(\mathbf{e})$ is a row of ones. This is achieved using
efficient multiplication with the sparse matrix created in the antenna-to-grid
mapping process. Unless $\widetilde{W}(\mathbf{r})$ includes time-dependent
effects of the ionosphere or the instrument, it needs to be computed just once
for the entire observation. However, the gridding of electric fields must be
computed at every readout of the electric field spectra. Thus,
\begin{align}
  \widetilde{E}(\mathbf{r}) &= \mathbf{M}(\mathbf{r},\mathbf{e})\,E(\mathbf{e}),
\end{align}
where, $E(\mathbf{e})$ denotes the spectra of measured antenna electric fields.

@@

@@

We use the EPIC simulator to generate electric field samples from a sky model. In
our example, we use 16 frequency channels each of width $\delta f=$100~kHz, 10
point sources of random flux densities at random locations. The number of
timestamps in one integration cycle was kept at four where each A/D timeseries is
$1/\delta f=10\,\mu$s long.

Show examples using simulations

Discuss PSF differences due to slight differences arising out of gridding

Apply it on LWA data

>>>>

The antenna array aperture illumination over the entire grid,
$\widetilde{W}(\mathbf{r})$, is obtained by a projection of the individual
antenna aperture illuminations:
\begin{align}\label{eqn:gridding-convolution}
  \widetilde{W}(\mathbf{r}) &= \sum_a \widetilde{W}_a(\mathbf{r}-\mathbf{r}_a) \\
                            &= \mathbf{M}(\mathbf{r},\mathbf{e})\,\mathcal{I}(\mathbf{e}),
\end{align}
where, $\mathcal{I}(\mathbf{e})$ is a row of ones. This is achieved using
efficient multiplication with the sparse matrix created in the antenna-to-grid
mapping process. Unless $\widetilde{W}(\mathbf{r})$ includes time-dependent
effects of the ionosphere or the instrument, it needs to be computed just once
for the entire observation. However, the gridding of electric fields must be
computed at every readout of the electric field spectra. Thus,
\begin{align}
  \widetilde{E}(\mathbf{r}) &= \mathbf{M}(\mathbf{r},\mathbf{e})\,E(\mathbf{e}),
\end{align}
where, $E(\mathbf{e})$ denotes the spectra of measured antenna electric fields.

@@

\subsection{Dealing with antenna auto-correlations}\label{sec:rm-autocorr}

The squaring operation under MOFF imaging in the image plane introduces antenna 
auto-correlations around the zero spacing in the $uv$-plane which are absent in 
traditional visibility-based imaging. In order for a robust comparison, these 
auto-correlations have to be removed from both the $uv$ and image planes.

The shape and extent of these auto-correlations can be estimated from the 
antenna aperture illumination pattern which is a quantity already available 
from the gridding step. Figure~\ref{fig:autocorr_wts_PB} shows the estimated 
weights from antenna auto-correlations in the $uv$-plane (left) and the 
corresponding response in the image plane (right). The latter is simply the 
directional antenna power response.

\begin{figure}
  \includegraphics[width=\columnwidth]{autocorr_uvwts_pbeam.png}
  \caption{The auto-correlation of weights of a square shaped antenna aperture
    in the $uv$ plane (left) and the corresponding directional antenna power 
    response on the sky (right) in coordinates specified by direction cosines. 
    The antenna auto-correlation weights are normalized to a sum
    of unity yielding a peak response of unity in the antenna's directional
    power pattern on the sky. The color scale for the directional power 
    pattern is logarithmic. The black circle indicates the sky horizon and
    values beyond it are not physical and hence ignored.}
  \label{fig:autocorr_wts_PB}
\end{figure}

We inverse Fourier transform the squared 
images and beams back to the $uv$ plane and subtract the estimated 
auto-correlation kernel scaled to the peak value centered at the zero spacing 
pixel. The final averaged image is obtained by Fourier transforming the $uv$ 
plane data and weights with the auto-correlations subtracted to the image plane. 
These images are now comparable to those obtained from visibility-based imaging. 
This step of removing auto-correlations is required to be performed only once 
per integration timescale and does not add significant cost to the full operation.

@@

We use the EPIC simulator to generate electric field samples from a sky model. In
our simulations, we use 64 frequency channels each of width $\delta f = 40$~kHz, 
10 point sources of flux densities 10~Jy at random locations. The number of
timestamps integrated in one integration cycle was kept at eight where each A/D 
timeseries is $1/\delta f=25\,\mu$s long. We use the MWA array layout 
\citep{bea12} for demonstration. Only the inner 51 tiles within a square bounding
box of 150~m on each side were used. We assumed all tiles are identical and have 
a square shaped electric field illumination footprint 4.4~m on each side.

Figure~\ref{fig:MOFF-FX-image} shows the dirty images (top) and synthesized beams
(bottom) obtained with antenna-based MOFF and FX visibility-based imaging 
algorithms packaged in EPIC. The antenna auto-correlations that correspond to 
zero spacing have been removed from the MOFF image and the corresponding 
synthesized beam. The sky positions of the simulated sources are indicated by 
solid black circles. The reconstructed sky image has the simulated sources at the
expected sky positions in either case. Both algorithms result in images and 
synthesized beams that are well matched with each other. Their fluxes are 
modulated by a multiplicative power pattern corresponding to that of a uniform
square aperture.

\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_FX_image_comparison.eps}
  \caption{Dirty images (top) and synthesized beams (bottom) obtained from 
    simulated data using antenna-based MOFF algorithm (left) and FX 
    visibility-based software holography (right). The solid black circles in the
    top panels indicate the simulated source positions. The antenna 
    auto-correlations at zero-spacing have been removed from the MOFF images. 
    The images in either case reconstruct the sources at the right locations 
    with the fluxes expected after multiplication by the antenna power pattern. 
    The synthesized beams from the two algorithms are well matches in size and
    shape. The overall modulation by the power pattern is seen clearly in both
    images.}
  \label{fig:MOFF-FX-image}
\end{figure}

We investigate the two imaging algorithms for differences from the point of 
view of the quality of their outputs. We begin by comparing the gridded 
cross-correlation weights in the $uv$-plane. In MOFF imaging, weights 
from antenna auto-correlations have been removed as described in 
\S\ref{sec:rm-autocorr}.

Figure~\ref{fig:MOFF-FX-uvwts} shows the cross-correlation weights obtained 
with MOFF imaging (left) and visibility-based imaging (right). The first 
notable difference is in the weights around zero spacing. Though both show a
dominant void around zero-spacing, the void obtained with MOFF algorithm shows 
many pixels with non-zero weights. In contrast, the zero-spacing void from 
traditional imaging consists of predominantly zero-valued pixels. The gridding
process in the former involves rounding the antenna footprint to the nearest 
grid pixel. Depending on the exact location of the center of the antenna 
relative to the grid, the grid pixels that receive contribution from an
antenna may be a pixel narrower along one or both axes relative to that from
another identical antenna but with a different center location relative to the
grid. The resulting auto-correlation footprint will also not necessarily be 
identical. Hence, using a single expected footprint for subtraction will
typically leave some residuals behind as seen in the void region in the 
weights obtained with MOFF imaging.

These residuals can be mitigated by: 
\begin{enumerate}
  \item making the grid spacing finer which makes the rounding error less 
    susceptible to the location of the antenna center relative to the grid, 
    and
  \item subtracting each auto-correlation of antenna weights separately by 
    using the shape and extent of the footprint appropriate for that specific
    antenna aperture. 
\end{enumerate}
The latter is the most general solution applicable especially in the case of 
heterogeneous antenna arrays.

The other notable difference is that weights outside the zero-spacing void in
some regions are different from each other at the few percent level though the 
sum of weights in these regions are identical. For instance, note the 
difference in weights at $(u,v)\approx (2,-10)$. This is found to arise 
because with MOFF imaging the antenna locations were rounded to the nearest 
grid pixel whereas in visibility-based imaging the baseline locations 
(difference between antenna locations) were rounded to the nearest grid pixel. 
Since rounding is not a commutative operation (i.e. rounding of the difference 
value is not necessarily equal to the difference of the rounded values), the 
gridding operation in the two cases introduces rounding errors in the placement 
of antenna aperture and cross-correlated aperture weights. Each antenna 
aperture weight projected to the nearest grid location can be displaced by 
$\sim 1$--2 pixels. This effect can be mitigated only by making the grid 
spacing finer.

\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_FX_comparison_uvwts_test_aperture_zero_spacing_removed.png}
  \caption{Dirty images (top) and synthesized beams (bottom) obtained from 
    simulated data using antenna-based MOFF algorithm (left) and FX 
    visibility-based software holography (right). The solid black circles in the
    top panels indicate the simulated source positions. The antenna 
    auto-correlations at zero-spacing have been removed from the MOFF images. 
    The images in either case reconstruct the sources at the right locations 
    with the fluxes expected after multiplication by the antenna power pattern. 
    The synthesized beams from the two algorithms are well matches in size and
    shape. The overall modulation by the power pattern is seen clearly in both
    images.}
  \label{fig:MOFF-FX-uvwts}
\end{figure}

We study the effect of the differences in gridded weights on the image plane. 
The synthesized beam can be written as:
\begin{align}\label{eqn:syn-beam}
  B(\hat{\mathbf{s}}) &= \left| \sum_\mathbf{r} \widetilde{W}(\mathbf{r})\,e^{-i\frac{2\pi}{\lambda}(\mathbf{r}\cdot\hat{\mathbf{s}})}\right|^2, \\
  \textrm{or}, \quad B(l,m) &= \left| \sum_{r_x,r_y} \widetilde{W}(r_x,r_y)\,e^{-i\frac{2\pi}{\lambda}(r_x l + r_y m)}\right|^2 
\end{align}
where, $\hat{\mathbf{s}}\equiv (l,m)$, $\mathbf{r}\equiv (r_x,r_y)$ and 
$\widetilde{W}(\mathbf{r})$ is given by equation~\ref{eqn:gridding-convolution}.
If the errors in the determination of $r_x$ and $r_y$ are denoted by 
$\delta r_x$ and $\delta r_y$ respectively, the error on $B(l,m)$ can be 
estimated using a simple propagation of errors, where,
\begin{align}
  \frac{\partial{B}(l,m)}{\partial{r_x}} &= e^{-i\frac{2\pi}{\lambda}\mathbf{r}\cdot\hat{\mathbf{s}}}\left[\frac{\partial{\widetilde{W}(r_x,r_y)}}{\partial{r_x}} - i\frac{2\pi l}{\lambda}\widetilde{W}(\mathbf{r})\right] \label{eqn:syn-beam-pdev1}\\
  &\qquad\qquad \times \sum_{\mathbf{r}^\prime} \widetilde{W}^\star(\mathbf{r}^\prime)\,e^{+i\frac{2\pi}{\lambda}\mathbf{r}^\prime\cdot \hat{\mathbf{s}}} \\
  &= e^{-i\frac{2\pi}{\lambda}\mathbf{r}\cdot\hat{\mathbf{s}}}\left[\frac{\partial{\widetilde{W}(r_x,r_y)}}{\partial{r_x}} - i\frac{2\pi l}{\lambda}\widetilde{W}(\mathbf{r})\right]\,W^\star(\hat{\mathbf{s}}), \\
  \textrm{and},\frac{\partial{B}(l,m)}{\partial{r_y}} &= e^{-i\frac{2\pi}{\lambda}\mathbf{r}\cdot\hat{\mathbf{s}}}\left[\frac{\partial{\widetilde{W}(r_x,r_y)}}{\partial{r_x}} - i\frac{2\pi m}{\lambda}\widetilde{W}(\mathbf{r})\right]\,W^\star(\hat{\mathbf{s}}). \label{eqn:syn-beam-pdev2}
\end{align}
It must be noted that although we start with a squaring operation in 
equation~\ref{eqn:syn-beam}, in equations~\ref{eqn:syn-beam-pdev1} through 
\ref{eqn:syn-beam-pdev2}, the summations are performed only when 
$\mathbf{r}\ne \mathbf{r}^\prime$ and therefore do not include the 
auto-correlations of the antenna weights. The error expected in $B(l,m)$ is:
\begin{align}
  \left|\delta B(l,m)\right|^2 &= \sum_{r_x}\left|\frac{\partial{B}(l,m)}{\partial{r_x}}\right|^2 \delta r_x^2 + \sum_{r_y}\left|\frac{\partial{B}(l,m)}{\partial{r_y}}\right|^2 \delta r_y^2 \\
  &= \delta r_x^2 \sum_{r_x}\left|\frac{\partial{\widetilde{W}(r_x,r_y)}}{\partial{r_x}} - i\frac{2\pi l}{\lambda}\widetilde{W}(\mathbf{r})\right|^2\,\left|W(\hat{\mathbf{s}})\right|^2 \nonumber\\
  &\qquad + \delta r_y^2 \sum_{r_y}\left|\frac{\partial{\widetilde{W}(r_x,r_y)}}{\partial{r_y}} - i\frac{2\pi m}{\lambda}\widetilde{W}(\mathbf{r})\right|^2\,\left|W(\hat{\mathbf{s}})\right|^2
\end{align}
For the current simulations, 
\begin{align}
  \delta r_x = \delta r_y &\simeq 1, \\
  \widetilde{W}(r_x,r_y) &= \{0, \textrm{constant}\}, \\
  \textrm{and},\qquad \frac{\partial{\widetilde{W}(r_x,r_y)}}{\partial{r_x}} &= \frac{\partial{\widetilde{W}(r_x,r_y)}}{\partial{r_y}} = 0.
\end{align}
Hence,
\begin{align}
  \left|\delta B(l,m)\right|^2 &\simeq \left(\frac{2\pi}{\lambda}\right)^2\left(l^2+m^2\right)\,B(\hat{\mathbf{s}})\,\sum_\mathbf{r}\,\left|\widetilde{W}(\mathbf{r})\right|^2, \\
\textrm{or},\qquad \left|\delta B(l,m)\right| &\simeq \left(\frac{2\pi}{\lambda}\right)\sqrt{l^2+m^2}\,\left|W(\hat{\mathbf{s}})\right|\,\left(\sum_\mathbf{r}\,\left|\widetilde{W}(\mathbf{r})\right|^2\right)^{1/2}, \label{eqn:syn-beam-error}
\end{align}
where, $B(\hat{\mathbf{s}}) \equiv \left|W(\hat{\mathbf{s}})\right|^2$. Thus, 
the error on the synthesized beam is proportional to the antenna power pattern
and increases linearly outward in radial bins of direction cosines.

Figure~\ref{fig:psf-diff} shows the difference between the synthesized beams 
obtained with the two methods. A difference map between the two synthesized 
beams is shown on top. The amplitude of the difference appears to be 
modulated by the directional voltage response of the antenna. At the bottom, 
in radial bins, the rms of the synthesized beam (gray) and the rms of the 
difference map (black) are plotted in percentage units relative to the peak 
(to be read using the axis on the left side of the plot). The antenna power 
pattern (red; to be read using the scale on the right) is plotted for reference.

\begin{figure}
  \includegraphics[width=\columnwidth]{diff_psf_MOFF-FX_test_aperture.png}
  \caption{Map of difference between the synthesized beams obtained with the 
    two methods (top) and radial statistics of the synthesized beams and 
    their differences (bottom). The maximum difference is of the order of a 
    few percent. The difference appears to be modulated in amplitude by the 
    power pattern of the antenna. }
  \label{fig:psf-diff}
\end{figure}

The synthesized beam rms is proportional to the antenna power pattern as 
expected from a point spread function of uncorrected for the antenna power 
pattern. A notable observation is that the rms of the difference of
the synthesized beams seems proportional to the directional antenna voltage
pattern and also increases gradually with distance from the center as seen 
from the diverging amplitudes between the black and red curves as predicted
in equation~\ref{eqn:syn-beam-error}. The rms of the difference appears to 
be bounded by the actual rms of the synthesized beam indicating that these
rounding errors in gridding do not affect the statistics of the images or
the synthesized beams.
